---
title: "Practice Problems"
subtitle: Nonparametric Stats
format: html
editor: visual
---

## Problem 1

(Nursery School Intervention). This dataset is drawn from
a study discussed by Siegel (1956). It involves eight pairs of identical twins who are of nursery school age. In the study, for each pair, one is randomly selected to attend nursery school while the other remains at home. At the end of the study period, all 16 children are given the same social awareness test. 

For each pair, the response of interest is:

```{r}
school = c(82,69,73,43,58,56,76,65)
home  = c(63,42,74,37,51,43,80,62)
```

Test the hypothesis that the nursery school twin has higher social awareness test score compared to the home school twin.


1. Use the Sign test
2. Use the Wilxocon signed ranks test


## Problem 2

This question is based on the case control study a cancer (Breslow et al. 1980). These data are available in the `datasets` package. 

```{r}
library(datasets)
head(esoph)
```

We want test the hypothesis that alcohol consumption is the same in the two groups, using as our dataset a sample of cases and controls.



